\chapter{Acknowledgements}

First and foremost I have to thank my advisors, Drs. Diane Re and Marianthi-Anna Kioumourtzoglou. Diane, thank you for giving me a place in your lab and your patience over the last five years. Your ongoing support in my research, training, and professional development has provided me with growth and learning. Marianthi, you have been an amazing professional, scientific, and personal mentor to me for the past 3 years. Thank you for admitting me in your lab halfway through my PhD, for your patience with my learning and for all your encouragement. I would not have been able to produce this body of work without support and guidance from both of you.

This work would also not have been possible without the guidance and advice of my committee members. Drs. Marc Weisskopf, Michael Aschner, Ana Navas-Acien, and Greg Freyer, thank you for your feedback and research advice throughout these projects. This work has greatly benefited from your critical comments during my committee meetings. Greg Freyer, thank you for always welcoming me into your office, listening to my concerns, and making me feel like you had my back. I truly appreciate your commitment towards education and the success of students. 

I would also like to thank Dr. Amelia K. Boehme for allowing me to work with the SPARCS data and to Drs. Randall V. Martin and Aaron V. Donkelaar for providing us with the air pollution data. The research work presented in this dissertation would not have been possible without their assistance. 

I would also like to thank Ana Abraido-Lanza and the IMSD program for funding the first two years of my PhD and for creating a safe and welcoming space to talk about the needs and issues faced in academia by people of color. Thank you also to the PRIME program for funding the last two years of my PhD and to the Mixtures Research team from whom I learned so much. Particularly, thanks to Dr. Jeff Goldsmith for teaching me to code and Lizzy Gibson for working with me through my first epidemiology project. 

Thank you also to Dr. Joe Graziano for the mentorship and to my long time mentor Dr. Alysson Muotri. Thank you Alysson for seeing my potential and nourishing my learning. You inspired my love for science, got me a job after college (in one of the most prestigious neuroscience labs), and supported me throughout the PhD application process --- I certainly would not be where I am without your support and teachings.   

Thank you also to my fellow students at Columbia University, current and former, for being such amazing colleagues, for your teachings, friendship, and smiles. This program would not have been the same without Dan Carrion, Alex Heaney, Anne Bozack, Richard Remigio, Sebastian Roland, 
Jenni Shearston, Ahlam Abuawad, Sarah Kramer, Mike He, and Roheeni Saxena. 

Lastly, on a more personal note, thank you to my family and friends for your endless love. Mami and Papi, thank you for teaching me diligence and courage --- I would not be here if it was not for you. Thanks to Daniel Hnatt for distracting me from work and being an amazing running partner, to Antonio Fernandez, Emily Johnson, Faith Umoh, Lauren Seibert, Rachel Sullivan, Thalia Aguirre, and the Appalachian Mountain Club Young Members Chapter who helped me balance my city and PhD life with unforgettable nature adventures. Finally, thank you Aaron Owji for being there for me through the highs and lows, for your endless encouragement and support, for proofreading all of my writing, for being my travel partner and for always believing in me. 

